Gammapy: A Python package for gamma-ray astronomy

### 2022-10-24 src/ms-review.tex
Update src/ms-review.tex
Co-authored-by: Cosimo Nigro <cosimonigro2@gmail.com>
Address more comments

\newcommand{\hess}{H.E.S.S.\xspace}
\newcommand{\hawc}{HAWC\xspace}
\newcommand{\veritas}{VERITAS\xspace}
\newcommand{\magic}{MAGIC\xspace}
\newcommand{\iact}{IACT\xspace}
\newcommand{\iacts}{IACTs\xspace}
\newcommand{\cta}{CTA\xspace}
\newcommand{\swgo}{SWGO\xspace}
\newcommand{\irf}{IRF\xspace}
\newcommand{\irfs}{IRFs\xspace}
\newcommand{\fermi}{\textit{Fermi}-LAT\xspace}
\newcommand{\gammaray}{$\gamma$-ray\xspace}
\newcommand{\gammarays}{$\gamma$-rays\xspace}
\newcommand{\gadf}{GADF\xspace}
\newcommand{\milagro}{MILAGRO\xspace}
\newcommand{\github}{GitHub\xspace}

@@

\abstract{
	% TODO: the feature are still to detailed and partly copy and pasted form the ICRS abstract...avoid this
	% context heading (optional)
	% {} leave it empty if necessary
	%{Gammapy context:}
	Historically, the data and the analysis software of \gammaray astronomy experiments are
	proprietary to the collaborations operating them. The next generation of \gammaray instruments, as the Cherenkov Telescope Array
	(CTA) will be operated as open observatories, with public
	data. This introduces the need to have a high-level analysis software available to the astronomical community. In
	this article we present the first major of \gammapy, the Long-term Support (LTS) v1.0, a
	community-developed open-source Python package for \gammaray astronomy. We
	present its general design and provide an
	overview of the analysis methods and features it implements. Starting from
	event lists and a description of the specific instrument response functions
	(IRF) stored in open FITS based data formats, \gammapy implements . Thereby it
	handles the dependency of the IRFs with time, energy as well as position on the
	sky. It offers a variety of background estimation methods for spectral, spatial
	and spectro-morphological analysis. Counts, background and IRFs data are
	bundled in datasets and can be serialised, rebinned and stacked. \gammapy
	supports to model binned data using Poisson
	maximum likelihood fitting. It comes with built-in spectral, spatial and
	temporal models as well as support for custom user models, to model e.g.,
	energy dependent morphology of \gammaray sources. Multiple datasets can be
	combined in a joint-likelihood approach to either handle time dependent IRFs,
	different classes of events or combination of data from multiple instruments.
	Gammapy also implements methods to estimate flux points, including likelihood
	profiles per energy bin, light curves as well as flux and signficance maps in
	energy bins.  We further describe the general
	development approach and how \gammapy integrates into ecosystem of other
	scientific and astronomical Python packages. We also present analysis examples
	with simulated CTA data and provide results of scientific validation analyses
	using data of existing instruments such as \hess and \fermi.

@@

%% Context
Ground based \gammaray astronomy has been historically structured
by experiments run by independent collaborations relying
on their own proprietary data and analysis software developed as part of the
instrument. While this model has been very successful so far, it does not
permit easy combination of data from several instruments and is therefore
a limitation for interoperability of existing facilities. Especially
because the different detection techniques have complementary
properties.

@@

%% Context : formats
In practice, the data reduction workflow of all \gammaray observatories
is remarkably similar. After data calibration, shower events are reconstructed and
gamma/hadron separation is applied to build lists of \gammaray like events.
The latter are then used to derive scientific results, such as spectra, sky maps
or light curves, taking into account the specific instrument response functions (IRF).
The information in this high data level is independent on
the data reduction, and eventually of the detection technique. This implies,
for example, that data from IACT and particle samplers can be represented
within the same model. The efforts to prototype a format usable by various instruments
converged in the so-called \textit{Data Format for \gammaray Astronomy}
initiative~\citep{gadf_proc, gadf_universe}, abbreviated in
\texttt{gamma-astro-data-format} (\gadf). The latter proposes prototypical
specifications to produce files based on the flexible image transport system
(FITS) format~\citep{fits} encapsulating this high-level information. This is
realized by storing a list of gamma-like events with their measured quantities
(energy, direction, arrival time) and a parametrisation of the response of the
system.

@@

The other sequence, referred to as high-level analysis, deals with the extraction of physical
quantities related to \gammaray sources and the production of high-level products such as spectra,
lightcurves and catalogs. The methods and tools applied here are more generic and are broadly
shared across the field. They also frequently imply joint analysis of multi-instrument data.
To extract physically relevant information, the measured data are usually compared to a
model of the expected \gammaray emitters in the instrument field-of-view using statistical
techniques such as maximum likelihood.

@@

The models follow a naming scheme which contains the category as a suffix to
the class name.
The spectral models include a special class of normed models, named using the NormSpectralModel suffix.
normalisation. These spectral models feature a dimension-less norm parameter
instead of a physical amplitude parameter. They
must be used along with another spectral model, as a multiplicative correction
factor according to their spectral shape. They can be typically used for
adjusting template based models, or adding a EBL correction to some analytic
model. The analytic spatial models are all normalized such as they integrate to
unity over the sky but the template spatial models may not, so in that special
case they have to be combined with a \code{NormSpectralModel}.

@@

In the \code{gammapy.astro.source} sub-module, dedicated classes exist for modeling
galactic \gammaray sources according to simplified physical models, e.g. SuperNova Remnant (SNR) evolution
models \citep{1950RSPSA.201..159T, 1999ApJS..120..299T}, evolution of Pulsar Wind Nebula (PWN) during the
free expansion phase \citep{2006ARA&A..44...17G} or computation
of physical parameters in a pulsar assumed to be a simple dipole.

@@

\subsection{Temporal Analysis}
\label{ssec:temporal-analysis}
A common use case in most astrophysical scenarios is to study the temporal
variability of a source. The most basic way to do this is to construct a
light curve, i.e., the flux of a source in each given time bin. In \gammapy, this
is done by using the \code{LightCurveEstimator} that fits the normalisation of a
source in each energy band per observation, keeping other parameters constant.
For custom time binning, an observation needs to be split into finer time bins using
the \code{Observation.select\_time} method. Figure~\ref{fig:hess_lightcurve_pks}
shows the light curve of the blazar PKS~2155-304 in different energy bands as
observed by the \hess telescope during an exceptional flare on the night of
July 29 - 30, 2006~\cite{2009A&A...502..749A}. The data are publicly available 
as a part of the HESS-DL3-DR1~\cite{HESS-DL3-DR1}, and shipped with
\verb"GAMMAPY_DATA". Each observation is first split into 10 min smaller
observations, and spectra extracted for each of these within a 0.11\textdegree radius
around the source. A \code{PowerLawSpectralModel} is fit to all the datasets, leading
to a reconstructed index of $3.54 \pm 0.02$. With this adjusted spectral model
the \code{LightCurveEstimator} runs directly for two energy bands, 0.5 - 1.5
TeV, and 1.5 - 20 TeV, respectively.
%
\begin{figure*}[t]
    \sidecaption
	\includegraphics[width=0.6666\textwidth]{figures/hess_lightcurve_pks.pdf}
	\caption{
        Binned light curves in two different energy bands for the source
        PKS~2155-304 in two energy bands, (500 GeV - 1.5 TeV, and 1.5 TeV to 20 TeV)
        as observed by the \hess telescopes in 2006. The coloured markers
        show the flux points in the different energy bands. The horizontal
        error illustrates the width of the time bin of 10~min. The vertical
        error bars show the associated assymetrical flux errors. The marker
        is set to the center of the time bin.
    }
    \label{fig:hess_lightcurve_pks}
\end{figure*}
%
The obtained flux points can be analytically modelled using the available or
user-implemented temporal models. Alternatively, instead of  extracting a
light curve, it is also possible to directly fit temporal models to the reduced
datasets. By associating an appropriate \code{SkyModel}, consisting of both temporal
and spectral components, or using custom temporal models with spectroscopic
variability, to each dataset, a joint fit across the datasets will directly
return the best fit temporal and spectral parameters.

@@

\subsection{Broadband SED Modeling}
\label{ssec:broadband-sed-modeling}
By combining \gammapy with astrophysical modelling codes, users can also fit
astrophysical spectral models to \gammaray data. In \gammaray
astronomy one typically observes two radiation production
mechanisms, the so-called hadronic and leptonic scenarios.
There are mutiple Python packages that are able to model
the \gammaray emission, given a physical scenario. Among those
packages are  \cite{agnpy}, \cite{naima}, \cite{jetset} or \cite{gamera}.
Tyically those emission models predict broadband emission from
radio waves up to the very high-energy \gammaray range.
By relying on the multiple dataset types in \gammapy those
data can be combined to constrain such a broadband emission model.
\gammapy provides a built-in \code{NaimaSpectralModel} that allows
users to wrap a given astrophysical emission model from the
naima package and fit it directly to \gammaray data.

@@

The Jupyter notebooks tutorials are stored stripped of their output cells,
which greatly helps in identifying differences in the contribution review.
During the CI these are executed, the output cells parsed and, in case an error
is found in one of the output cells then validation fails. Code formatting of
the input cells is also done, parsing their content and using a particular
syntax to provide links to the API documentation pages. The resulting web
published tutorials also provide links to playground spaces in
"myBinder"~\citep{project_jupyter-proc-scipy-2018}, where they may be executed
on-line in versioned virtual environments hosted in the myBinder
infrastructure. Users may also play with the tutorials locally in their
laptops. They can download a specific version of the tutorials together with
the associated datasets needed and the specific conda computing environment,
using the \textit{gammapy download} command.

@@

Around the core Python package a large diverse community of
users and contributors has developed. With regular developer meetings,
coding sprints and in-person user tutorials at relevant conferences
and collaboration meetings, the community has constantly grown.
So far \gammapy has seen ~80 contributors from ~10 different countries.
With typically ~10 regular contributors at any given time of the
project, the code base has constantly grown its range of features
and improved its code quality. With \gammapy being selected as the base library
for the future science tools for CTA, we expect the community to grow
even further, providing a stable perspective for further usage,
development and maintenance of the project. Besides the future use
by the CTA community \gammapy has already
been used for analysis of data from the \hess and \magic instruments

>>>>

\newcommand{\hess}{H.E.S.S.\xspace}
\newcommand{\hawc}{HAWC\xspace}
\newcommand{\veritas}{VERITAS\xspace}
\newcommand{\magic}{MAGIC\xspace}
\newcommand{\astri}{ASTRI\xspace}
\newcommand{\iact}{IACT\xspace}
\newcommand{\iacts}{IACTs\xspace}
\newcommand{\cta}{CTA\xspace}
\newcommand{\swgo}{SWGO\xspace}
\newcommand{\irf}{IRF\xspace}
\newcommand{\irfs}{IRFs\xspace}
\newcommand{\fermi}{\textit{Fermi}-LAT\xspace}
\newcommand{\gammaray}{$\gamma$-ray\xspace}
\newcommand{\gammarays}{$\gamma$-rays\xspace}
\newcommand{\gadf}{GADF\xspace}
\newcommand{\milagro}{MILAGRO\xspace}
\newcommand{\github}{GitHub\xspace}

@@

\abstract{
	% TODO: the feature are still to detailed and partly copy and pasted form the ICRS abstract...avoid this
	% context heading (optional)
	% {} leave it empty if necessary
	%{Gammapy context:}
	Historically, the data and the analysis software of \gammaray astronomy experiments are
	proprietary to the collaborations operating them. The next generation of \gammaray instruments, as the Cherenkov Telescope Array
	(CTA) will be operated as open observatories, with public
	data. This introduces the need to have a high-level analysis software available to the astronomical community. In
	this article we present the first major of \gammapy, the Long-term Support (LTS) v1.0, a
	community-developed open-source Python package for \gammaray astronomy. We
	present its general design and provide an
	overview of the analysis methods and features it implements. Starting from
	event lists and a description of the specific instrument response functions
	(IRF) stored in open FITS based data formats, \gammapy implements the reduction
	of the input data and instrument response to binned WCS, HEALPix or region based
	data structures. Thereby it
	handles the dependency of the IRFs with time, energy as well as position on the
	sky. It offers a variety of background estimation methods for spectral, spatial
	and spectro-morphological analysis. Counts, background and IRFs data are
	bundled in datasets and can be serialised, rebinned and stacked. \gammapy
	supports to model binned data using Poisson
	maximum likelihood fitting. It comes with built-in spectral, spatial and
	temporal models as well as support for custom user models, to model e.g.,
	energy dependent morphology of \gammaray sources. Multiple datasets can be
	combined in a joint-likelihood approach to either handle time dependent IRFs,
	different classes of events or combination of data from multiple instruments.
	Gammapy also implements methods to estimate flux points, likelihood
	profiles per energy bin, light curves as well as flux and signficance maps in
	energy bins.  We further describe the general
	development approach and how \gammapy integrates into ecosystem of other
	scientific and astronomical Python packages. We also present analysis examples
	with simulated CTA data and provide results of scientific validation analyses
	using data of existing instruments such as \hess and \fermi.

@@

%% Context
Ground based \gammaray astronomy has been historically structured
by experiments run by independent collaborations relying
on their own proprietary data and analysis software developed as part of the
instrument. While this model has been very successful so far, it does not
permit easy combination of data from several instruments and therefore
limits the interoperability of existing facilities. This lack of
interoperability currently limits the full explotation of the
available \gammaray data, especially because the different detection
techniques have complementary properties in terms of covered energy rage,
dutu cycle and spatial resolution.

@@

%% Context : formats
In practice, the data reduction workflow of all \gammaray observatories
is remarkably similar. After data calibration, shower events are reconstructed and
gamma/hadron separation is applied to build lists of \gammaray like events.
The lists of \gammary events are then used to derive scientific results, such as spectra, sky maps
or light curves, taking into account the observatory specific instrument response functions (IRF).
One the data is present in the form of a list of events, the information is independent on
the data reduction process, and eventually of the detection technique. This implies,
for example, that data from IACT and particle samplers can be represented
within the same model. The efforts to prototype a format usable by various instruments
converged in the so-called \textit{Data Format for \gammaray Astronomy}
initiative~\citep{gadf_proc, gadf_universe}, abbreviated in
\texttt{gamma-astro-data-format} (\gadf). The latter proposes prototypical
specifications to produce files based on the flexible image transport system
(FITS) format~\citep{fits} encapsulating this high-level information. This is
realized by storing a list of gamma-like events with their measured quantities
(energy, direction, arrival time) and a parametrisation of the response of the
system.

@@

The other sequence, referred to as high-level analysis, deals with the extraction of physical
quantities related to \gammaray sources and the production of high-level products such as spectra,
light curves and catalogs. The methods and tools applied here are more generic and are broadly
shared across the field. They also frequently imply joint analysis of multi-instrument data.
To extract physically relevant information, the measured data are usually compared to a
model of the expected \gammaray emitters in the instrument field-of-view using statistical
techniques such as maximum likelihood.

@@

The models follow a naming scheme which contains the category as a suffix to
the class name.
The spectral models include a special class of normed models, named using the \code{NormSpectralModel} suffix.
These spectral models feature a dimension-less norm parameter
instead of an amplitude parameter with physical units. They
must be used along with another spectral model, as a multiplicative correction
factor according to their spectral shape. They can be typically used for
adjusting template based models, or adding taking into account
the absorbtion effect on 
\gammaray spectra caused by the extra-galactic background light (EBL) (\code{EBLAbsorptionNormSpectralModel})
model. \gammapy supports a variety of EBL absorbtion models, such as

\cite{Franceschini2008}, \cite{Finke2010} and \cite{Dominguez2011}.

The analytic spatial models are all normalized such as they integrate to
unity over the sky but the template spatial models may not, so in that special
case they have to be combined with a \code{NormSpectralModel}.

@@

In the \code{gammapy.astro.source} sub-module, dedicated classes exist for modeling
galactic \gammaray sources according to simplified physical models, e.g. Supernova Remnant (SNR) evolution
models \citep{1950RSPSA.201..159T, 1999ApJS..120..299T}, evolution of Pulsar Wind Nebula (PWN) during the
free expansion phase \citep{2006ARA&A..44...17G} or computation
of physical parameters of a pulsar using a simplified dipole spin-down model.

@@

\subsection{Temporal Analysis}
\label{ssec:temporal-analysis}
A common use case in most astrophysical scenarios is to study the temporal
variability of a source. The most basic way to do this is to construct a
light curve, i.e., the flux of a source in each given time bin. In \gammapy, this
is done by using the \code{LightCurveEstimator} that fits the normalisation of a
source in each time (and optionally energy) band per observation, keeping other parameters constant.
For custom time binning, an observation needs to be split into finer time bins using
the \code{Observation.select\_time} method. Figure~\ref{fig:hess_lightcurve_pks}
shows the light curve of the blazar PKS~2155-304 in different energy bands as
observed by the \hess telescope during an exceptional flare on the night of
July 29 - 30, 2006~\cite{2009A&A...502..749A}. The data are publicly available 
as a part of the HESS-DL3-DR1~\cite{HESS-DL3-DR1}, and shipped with
\verb"GAMMAPY_DATA". Each observation is first split into 10 min smaller
observations, and spectra extracted for each of these within a 0.11\textdegree radius
around the source. A \code{PowerLawSpectralModel} is fit to all the datasets, leading
to a reconstructed index of $3.54 \pm 0.02$. With this adjusted spectral model
the \code{LightCurveEstimator} runs directly for two energy bands, 0.5 - 1.5
TeV, and 1.5 - 20 TeV, respectively.
%
\begin{figure*}[t]
    \sidecaption
	\includegraphics[width=0.6666\textwidth]{figures/hess_lightcurve_pks.pdf}
	\caption{
        Binned light curves in two different energy bands for the source
        PKS~2155-304 in two energy bands, (500 GeV - 1.5 TeV, and 1.5 TeV to 20 TeV)
        as observed by the \hess telescopes in 2006. The coloured markers
        show the flux points in the different energy bands. The horizontal
        error illustrates the width of the time bin of 10~min. The vertical
        error bars show the associated assymetrical flux errors. The marker
        is set to the center of the time bin.
    }
    \label{fig:hess_lightcurve_pks}
\end{figure*}
%
The obtained flux points can be analytically modelled using the available or
user-implemented temporal models. Alternatively, instead of  extracting a
light curve, it is also possible to directly fit temporal models to the reduced
datasets. By associating an appropriate \code{SkyModel}, consisting of both temporal
and spectral components, or using custom temporal models with spectroscopic
variability, to each dataset, a joint fit across the datasets will directly
return the best fit temporal and spectral parameters.

@@

\subsection{Broadband SED Modeling}
\label{ssec:broadband-sed-modeling}
By combining \gammapy with astrophysical modelling codes, users can also fit
astrophysical spectral models to \gammaray data. In \gammaray
astronomy one typically observes two radiation production
mechanisms, the so-called hadronic and leptonic scenarios.
There are mutiple Python packages that are able to model
the \gammaray emission, given a physical scenario. Among those
packages are  \cite{agnpy}, \cite{naima}, \cite{jetset} or \cite{gamera}.
Tyically those emission models predict broadband emission from
radio, up to the very high-energy \gammaray range.
By relying on the multiple dataset types in \gammapy those
data can be combined to constrain such a broadband emission model.
\gammapy provides a built-in \code{NaimaSpectralModel} that allows
users to wrap a given astrophysical emission model from the
naima package and fit it directly to \gammaray data.

@@

The Jupyter notebooks tutorials are generated using the sphinx-gallery
package \citep{sphinx-gallery}.
The resulting web published tutorials also provide links to playground spaces in
"myBinder"~\citep{project_jupyter-proc-scipy-2018}, where they may be executed
on-line in versioned virtual environments hosted in the myBinder
infrastructure. Users may also play with the tutorials locally in their
laptops. They can download a specific version of the tutorials together with
the associated datasets needed and the specific conda computing environment,
using the \textit{gammapy download} command.

@@

Around the core Python package a large diverse community of
users and contributors has developed. With regular developer meetings,
coding sprints and in-person user tutorials at relevant conferences
and collaboration meetings, the community has constantly grown.
So far \gammapy has seen ~80 contributors from ~10 different countries.
With typically ~10 regular contributors at any given time of the
project, the code base has constantly grown its range of features
and improved its code quality. With \gammapy being selected as the base library
for the future science tools for CTA, we expect the community to grow
even further, providing a stable perspective for further usage,
development and maintenance of the project. Besides the future use
by the CTA community \gammapy has already
been used for analysis of data from the \hess, \magic
\hawc, \astri and \veritas instruments.
